Spike-and-Slab LASSO Generalized Additive Models and Scalable Algorithms for High-Dimensional Data Analysis

### 2022-05-16 Manuscript/SS_GAM.tex
* Remove useless markdowns * Correct smoothing penalty equation
remove revision highlights in red

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{*}
\newcommand{\pr}{\text{Pr}}
\newcommand{\repa}{\text{repa}}
\newcommand{\simiid}{\overset{\text{iid}}{\sim}}
\newcommand{\bg}[1]{\textcolor{red}{#1}}

@@

Much modern biomedical research, e.g., sequencing data analysis and
electronic health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are useful for statistical prediction but do not estimate parameters
relevant to the data generation process that one can draw inferences
from. In addition, how to generalize these \textcolor{red}{"black box"}
models to the high-dimensional setting remains unclear.

\textcolor{red}{For their straightforward interpretation and flexibility, nonparametric regression models serve as great alternatives to the "black-box" models in prediction and variable selection}.
Among those, generalized additive models (GAMs), proposed in the seminal
work of Hastie and Tibshirani \citep{Hastie1987}, grew to be one of the
most popular modeling tools. In a GAM, the response variable, which is
assumed to follow some exponential family distribution, can be modeled
with the summation of smooth functions. Nevertheless, the classical GAMs
cannot fulfill the increasing analytic demands for high-dimensional data
analysis.

@@

\textcolor{red}{We propose a novel Bayesian hierarchical generalized additive model (BHAM) for outcome prediction in the context of high-dimensional data analysis. Specifically, we incorporate smoothing penalties, derived from the smoothing spline literature \cite{Wood2017}, via reparameterization of smooth functions to avoid excessive shrinkage on the bases. Smoothing penalties were also previously used in the spike-and-slab GAM \cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}. We then impose a new two-part spike-and-slab LASSO prior to address the signal sparsity. In addition, a scalable optimization-based algorithm, EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary focus of this model is to improve prediction, the proposed model also provides utility in functional selection. Notably, the two-part prior that follows the effect hierarchy principle motivates a bi-level selection, rendering one of three possibilities for each predictor: no effect, only linear effect, or linear and nonlinear effects. The proposed model is implemented in a publicly available R package \texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.}

@@

\textcolor{red}{We assume the response variable, $Y$, follows an exponential family distribution with density function $f(y)$, mean $\mu$ and dispersion parameter $\phi$. The mean of the response variable can be modeled as the summation of smooth functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of predictors $\boldsymbol{x}$, written as 
\begin{equation}\label{eq:gam}
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation}
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. Given $n$ data points $\{y_i, \boldsymbol{x}_i\}^n_{i=1}$,}
the data distribution is expressed as \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi).\nonumber
\end{equation}

The basis function matrix, i.e.~the design matrix derived from the
smooth function \(B_j(x_j)\), is denoted \(\boldsymbol{X}_j\) for the
variable \(x_j\). The dimension of the design matrix depends on the
choice of the smooth function, and is denoted as \(K_j\) for \(x_j\).
\(\boldsymbol{\beta}_j\) denotes the basis function coefficients for the
\(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\).
\textcolor{red}{To note, the proposed model can include parametric forms of variables in the model, and hence considers general linear models and semiparametric regression models as special cases.}

\subsection{smooth function Reparameterization}

To encourage proper smoothing of each additive function, we adopt the
smoothing penalty from smoothing spline models\citep{Wood2017}. A
smoothing penalty is the quadratic norm of the basis coefficients and
allows different shrinkage on different bases, mathematically
\begin{equation}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,\nonumber
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix and \(\lambda_j\) denotes a smoothing parameter. A linear
function can be modeled as \(B_j(x_j) = x_j\) with the smoothing penalty
matrix \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).
\textcolor{red}{Unlike previous regularized methods that either ignore the smoothing penalty completely or restrain the smoothing penalty as a component of sparse penalty which leads to a more restrictive solution, we consider an additional mechanism in pair with the proposed prior (described in Section \ref{sec:method_prior}) to address the smoothness and sparsity in signals such that the locally adaptive nature of the smoothing penalty retains.}

Marra and Wood \citep{Marra2011} proposed a reparameterization procedure
\textcolor{red}{to factor the smoothing penalty into the design matrix of each smooth function}.
Given that the smoothing penalty matrix \(\boldsymbol{S}_j\) is
symmetric and positive semi-definite for the univariate smooth
functions, we eigendecompose the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smooth function. For the most popular smooth
function, cubic splines, the dimension of the linear space is one.
Hereafter, we focus on discussing a uni-dimensional linear space for
simplicity; however, it generalizes easily to the cases where the linear
space is multidimensional. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the nonlinear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive a uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
reparameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) now is
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of nonlinear
space.

@@

The family of spike-and-slab regression models is one of the most
commonly used models in high-dimensional data analysis for its utility
in outcome prediction and variable selection.
\textcolor{red}{Among all the spike-and-slab priors, the spike-and-slab LASSO (SSL) prior \cite{Rockova2018b, Rockova2018} is one of the most popular choices because it's highly scalable. The SSL prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
} The latent binary variable \(\gamma \in \{0,1\}\) indicates whether a
variable \(x\) is included in the model, while the dispersion parameters
\(s_0\) and \(s_1\) control the shrinkage of the coefficient. Given that
both double exponential distributions have a mean of 0 and the latent
indicator \(\gamma\) can only take the value of 0 or 1, the mixture
double exponential distribution can be formulated as one single double
exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, (1-\gamma)s_0 + \gamma s_1), 0 < s_0 < s_1.
\end{equation}
\textcolor{red}{Compared to other priors for high-dimensional data analysis, SSL has the following advantages. First of all, the SSL prior provides a locally adaptive shrinkage when estimating the coefficients. Secondly, the SSL prior encourages a sparse solution, making variable selection straightforward. Thirdly, the SSL prior motivates a scalable algorithm, the EM-CD algorithm, for model fitting, and hence is more feasible for high-dimensional data analysis. We defer to Bai et al. \cite{Bai2021Review} for a detailed discussion.}

@@

\textcolor{red}{The proposed two-part SSL prior, particularly the group SSL prior of the nonlinear space coefficients, differs from previous group SSL priors \cite{Tang2018, Tang2019}, as the proposed prior follows the effect hierarchy principle. Effect hierarchy refers to the principle that "lower-order effects are more likely to be active than higher-order effects" defined by Chipman\cite{chipman2006prior}. To implement, we consider the shared latent indicator of nonlinear coefficients $\gamma^*_j$}
depends on the value of the linear space latent indicator \(\gamma_j\),
while both latent indicators \(\gamma_j\) and \(\gamma^*_j\) follow a
Bernoulli distribution. While the probability of including the linear
effect is \(\theta_j\), the probability of including the nonlinear
effect is \(\textcolor{red}{\gamma_{j}}\theta_j\). \[
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^*| \textcolor{red}{\gamma_{j}, } \theta_j \sim Bin(1, \textcolor{red}{\gamma_{j}}\theta_j).
\end{aligned}
\]
\textcolor{red}{This is, when the linear effect is not selected, the probability of including the nonlinear effect drops from $\theta_j$ to 0. For computational convenience, we analytically integrate $\gamma_j$ out such that $\gamma_{j}^*| \theta_j \sim Bin(1, \theta_j^2)$ (see the derivation in the Supporting Information).
}

\textcolor{red}{To allow the shrinkage to self-adapt to the sparsity and smoothing pattern of the data, }we
further specify the parameter \(\theta_j\) follows a beta distribution
with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior to the binomial
distribution and hence provides some computation convenience.
\textcolor{red}{Having a prior distribution of $\theta_j$ enables the proposed prior to inherit the selective shrinkage property and self-adaptivity \cite{Bai2021Review} from the classical SSL prior. In other words, when a smooth function is significant, the coefficients of the smooth function escape the overall shrinkage and produce a more accurate estimate, particularly in pair with the smoothing penalty implicitly addressed via the reparameterization. Meanwhile, the hyper prior encourages information borrowing across coordinates and hence automatic adjust for different levels of sparsity.}
Hereafter, we refer to the Bayesian hierarchical generalized additive
models with the two-part spike-and-slab LASSO prior as BHAM, and
visually presented in Figure \ref{fig:SSprior}.

@@

\textcolor{red}{Despite the advantage to estimate posterior densities, using MCMC algorithms to fit the proposed model is computationaly prohibited and not feasible for high-dimensional data. Previous research shows the computation performance of MCMC algorithms for spike-and-slab models is bottlenecked for medium-sized data ($p$=25) \cite{George1997}, and substantially slows as $p$ increases modestly in the GAM context \cite{Scheipl2013}. Hence, we consider the optimization algorithms that focus on the maximum a posteriori estimates at the cost of posterior inference. Specifically, we
extend the EM-Coordinate Descent (EM-CD) algorithm to fit BHAMs. Similar to the EMVS algorithm \cite{Rockova2014a} for spike-and-slab models, the EM-CD algorithm is based on the expectation-maximization (EM) algorithm, integrating the Coordinate Descent algorithm in each iterative step to find the posterior mode. The EM-CD algorithm has been well adapted in generalized linear models \cite{Tang2017a}, Cox proportional hazards models \cite{Tang2017}, and their grouped counterparts \cite{Tang2018, Tang2019}. The EM-CD algorithm provides deterministic solutions, which becomes a popular property for reproducible research.}

@@

We use the EM algorithm to find the maximum a posteriori estimate of
\(\Theta\). This is, in the E-step, we calculate the expectation of
posterior density function of
\(\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\) with
respect to the latent indicators \(\boldsymbol{\gamma}\) conditioning on
the parameter values from previous iteration \(\Theta^{(t-1)}\), \[
E_{\boldsymbol{\gamma}|\Theta^{(t-1)}}\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) .
\] Hereafter, we use the shorthand notation
\(E(\cdot)\equiv E_{\boldsymbol{\gamma}|\Theta^{(t-1)}}(\cdot)\). In the
M-step, we find the parameters \(\Theta^{(t)}\) that maximize
\(E\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\).
\textcolor{red}{The parenthesized subscription $^{(t)}$ denotes the parameter estimation at the $t$th iteration.}
The E- and M- steps are iterated until the algorithm converges.

@@

Our proposed model, BHAM, requires two preset scale parameters (\(s_0\),
\(s_1\)). Hence, we need to find the optimal values for the scale
parameters such that the model reaches its best prediction performance
regarding criteria of preference. This would be achieved by constructing
a two-dimensional grid, consisting of different (\(s_0\), \(s_1\))
pairs. However, previous research suggests the value of slab scale
\(s_1\) has less impact on the final model and is recommended to be set
as a generally large value, e.g.~\(s_1 = 1\), that provides no or weak
shrinkage. \citep{Rockova2018} As a result, we focus on examining
different values of spike scale \(s_0\). Instead of the two-dimensional
grid, we consider a sequence of \(L\) decreasing values
\(\{s_0^l\}: 0 < s_0^1 < s_0^2 < \dots < s_0^L < s_1\). Increasing the
spike scale \(s_0\) tends to include more non-zero coefficients in the
model. A measure of preference calculated with cross-validations (CV),
e.g.~deviance
\textcolor{red}{(defined as model log-likelihood times -2, $-2\log f(\boldsymbol{y}|\boldsymbol{\hat\beta}, \hat\phi)$)},
area under the curve (AUC), mean squared error, can be used to
facilitate the selection of a final model. The procedure is similar to
the LASSO implementation in the widely used R package \texttt{glmnet},
which quickly fits LASSO models over a list of candidate values of
\(\lambda\) and gives a sequence of models for users to choose from.

@@

In this section, we apply the proposed model BHAM to analyze two
published metabolomics datasets where the outcomes are binary and
continuous respectively. We demonstrate the improved prediction
performance compared to the other Bayesian hierarchical additive model,
SB-GAM \citep{Bai2021}, while being computationally efficient (see Table
\ref{tab:time_real_data}).
\textcolor{red}{BHAM requires roughly 10\% of the computation time of SB-GAM to fit models.}

>>>>

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{*}
\newcommand{\pr}{\text{Pr}}
\newcommand{\repa}{\text{repa}}
\newcommand{\simiid}{\overset{\text{iid}}{\sim}}

@@

Much modern biomedical research, e.g., sequencing data analysis and
electronic health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are useful for statistical prediction but do not estimate parameters
relevant to the data generation process that one can draw inferences
from. In addition, how to generalize these ``black box'' models to the
high-dimensional setting remains unclear.

For their straightforward interpretation and flexibility, nonparametric
regression models serve as great alternatives to the ``black-box''
models in prediction and variable selection. Among those, generalized
additive models (GAMs), proposed in the seminal work of Hastie and
Tibshirani \citep{Hastie1987}, grew to be one of the most popular
modeling tools. In a GAM, the response variable, which is assumed to
follow some exponential family distribution, can be modeled with the
summation of smooth functions. Nevertheless, the classical GAMs cannot
fulfill the increasing analytic demands for high-dimensional data
analysis.

@@

We propose a novel Bayesian hierarchical generalized additive model
(BHAM) for outcome prediction in the context of high-dimensional data
analysis. Specifically, we incorporate smoothing penalties, derived from
the smoothing spline literature \cite{Wood2017}, via reparameterization
of smooth functions to avoid excessive shrinkage on the bases. Smoothing
penalties were also previously used in the spike-and-slab GAM
\cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}.
We then impose a new two-part spike-and-slab LASSO prior to address the
signal sparsity. In addition, a scalable optimization-based algorithm,
EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary
focus of this model is to improve prediction, the proposed model also
provides utility in functional selection. Notably, the two-part prior
that follows the effect hierarchy principle motivates a bi-level
selection, rendering one of three possibilities for each predictor: no
effect, only linear effect, or linear and nonlinear effects. The
proposed model is implemented in a publicly available R package
\texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.

@@

We assume the response variable, \(Y\), follows an exponential family
distribution with density function \(f(y)\), mean \(\mu\) and dispersion
parameter \(\phi\). The mean of the response variable can be modeled as
the summation of smooth functions, \(B_j(\cdot), j = 1, \dots, p\), of a
given \(p\)-dimensional vector of predictors \(\boldsymbol{x}\), written
as \begin{equation}\label{eq:gam}
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} where \(g^{-1}(\cdot)\) is the inverse of a monotonic
link function. Given \(n\) data points
\(\{y_i, \boldsymbol{x}_i\}^n_{i=1}\), the data distribution is
expressed as \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi).\nonumber
\end{equation}

The basis function matrix, i.e.~the design matrix derived from the
smooth function \(B_j(x_j)\), is denoted \(\boldsymbol{X}_j\) for the
variable \(x_j\). The dimension of the design matrix depends on the
choice of the smooth function, and is denoted as \(K_j\) for \(x_j\).
\(\boldsymbol{\beta}_j\) denotes the basis function coefficients for the
\(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, and hence considers general linear
models and semiparametric regression models as special cases.

\subsection{Smooth Function Reparameterization}

To encourage proper smoothing of each additive function, we adopt the
smoothing penalty from smoothing spline models\citep{Wood2017}. A
smoothing penalty is the quadratic norm of the basis coefficients and
allows different shrinkage on different bases, mathematically
\begin{equation}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)^2dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,\nonumber
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix and \(\lambda_j\) denotes a smoothing parameter. A linear
function can be modeled as \(B_j(x_j) = x_j\) with the smoothing penalty
matrix \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\). Unlike
previous regularized methods that either ignore the smoothing penalty
completely or restrain the smoothing penalty as a component of sparse
penalty which leads to a more restrictive solution, we consider an
additional mechanism in pair with the proposed prior (described in
Section \ref{sec:method_prior}) to address the smoothness and sparsity
in signals such that the locally adaptive nature of the smoothing
penalty retains.

Marra and Wood \citep{Marra2011} proposed a reparameterization procedure
to factor the smoothing penalty into the design matrix of each smooth
function. Given that the smoothing penalty matrix \(\boldsymbol{S}_j\)
is symmetric and positive semi-definite for the univariate smooth
functions, we eigendecompose the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smooth function. For the most popular smooth
function, cubic splines, the dimension of the linear space is one.
Hereafter, we focus on discussing a uni-dimensional linear space for
simplicity; however, it generalizes easily to the cases where the linear
space is multidimensional. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the nonlinear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive a uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
reparameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) now is
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of nonlinear
space.

@@

The family of spike-and-slab regression models is one of the most
commonly used models in high-dimensional data analysis for its utility
in outcome prediction and variable selection. Among all the
spike-and-slab priors, the spike-and-slab LASSO (SSL) prior
\cite{Rockova2018b, Rockova2018} is one of the most popular choices
because it's highly scalable. The SSL prior is composed of two double
exponential distributions with mean 0 and different dispersion
parameters, \(0 < s_0 < s_1\), mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} The latent binary variable \(\gamma \in \{0,1\}\)
indicates whether a variable \(x\) is included in the model, while the
dispersion parameters \(s_0\) and \(s_1\) control the shrinkage of the
coefficient. Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
one single double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, (1-\gamma)s_0 + \gamma s_1), 0 < s_0 < s_1.
\end{equation} Compared to other priors for high-dimensional data
analysis, SSL has the following advantages. First of all, the SSL prior
provides a locally adaptive shrinkage when estimating the coefficients.
Secondly, the SSL prior encourages a sparse solution, making variable
selection straightforward. Thirdly, the SSL prior motivates a scalable
algorithm, the EM-CD algorithm, for model fitting, and hence is more
feasible for high-dimensional data analysis. We defer to Bai et
al.~\cite{Bai2021Review} for a detailed discussion.

@@

The proposed two-part SSL prior, particularly the group SSL prior of the
nonlinear space coefficients, differs from previous group SSL priors
\cite{Tang2018, Tang2019}, as the proposed prior follows the effect
hierarchy principle. Effect hierarchy refers to the principle that
``lower-order effects are more likely to be active than higher-order
effects'' defined by Chipman\cite{chipman2006prior}. To implement, we
consider the shared latent indicator of nonlinear coefficients
\(\gamma^*_j\) depends on the value of the linear space latent indicator
\(\gamma_j\), while both latent indicators \(\gamma_j\) and
\(\gamma^*_j\) follow a Bernoulli distribution. While the probability of
including the linear effect is \(\theta_j\), the probability of
including the nonlinear effect is \(\gamma_{j}\theta_j\). \[
\begin{aligned}
&\gamma_{j} | \theta_j \sim Bin(1, \theta_j) & & 
&\gamma_{j}^*| \gamma_{j}, \theta_j \sim Bin(1, \gamma_{j}\theta_j).
\end{aligned}
\] This is, when the linear effect is not selected, the probability of
including the nonlinear effect drops from \(\theta_j\) to 0. For
computational convenience, we analytically integrate \(\gamma_j\) out
such that \(\gamma_{j}^*| \theta_j \sim Bin(1, \theta_j^2)\) (see the
derivation in the Supporting Information).

To allow the shrinkage to self-adapt to the sparsity and smoothing
pattern of the data, we further specify the parameter \(\theta_j\)
follows a beta distribution with given shape parameters \(a\) and \(b\),
\[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior to the binomial
distribution and hence provides some computation convenience. Having a
prior distribution of \(\theta_j\) enables the proposed prior to inherit
the selective shrinkage property and self-adaptivity
\cite{Bai2021Review} from the classical SSL prior. In other words, when
a smooth function is significant, the coefficients of the smooth
function escape the overall shrinkage and produce a more accurate
estimate, particularly in pair with the smoothing penalty implicitly
addressed via the reparameterization. Meanwhile, the hyper prior
encourages information borrowing across coordinates and hence automatic
adjust for different levels of sparsity. Hereafter, we refer to the
Bayesian hierarchical generalized additive models with the two-part
spike-and-slab LASSO prior as BHAM, and visually presented in Figure
\ref{fig:SSprior}.

@@

Despite the advantage to estimate posterior densities, using MCMC
algorithms to fit the proposed model is computationaly prohibited and
not feasible for high-dimensional data. Previous research shows the
computation performance of MCMC algorithms for spike-and-slab models is
bottlenecked for medium-sized data (\(p\)=25) \cite{George1997}, and
substantially slows as \(p\) increases modestly in the GAM context
\cite{Scheipl2013}. Hence, we consider the optimization algorithms that
focus on the maximum a posteriori estimates at the cost of posterior
inference. Specifically, we extend the EM-Coordinate Descent (EM-CD)
algorithm to fit BHAMs. Similar to the EMVS algorithm
\cite{Rockova2014a} for spike-and-slab models, the EM-CD algorithm is
based on the expectation-maximization (EM) algorithm, integrating the
Coordinate Descent algorithm in each iterative step to find the
posterior mode. The EM-CD algorithm has been well adapted in generalized
linear models \cite{Tang2017a}, Cox proportional hazards models
\cite{Tang2017}, and their grouped counterparts
\cite{Tang2018, Tang2019}. The EM-CD algorithm provides deterministic
solutions, which becomes a popular property for reproducible research.

@@

We use the EM algorithm to find the maximum a posteriori estimate of
\(\Theta\). This is, in the E-step, we calculate the expectation of
posterior density function of
\(\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\) with
respect to the latent indicators \(\boldsymbol{\gamma}\) conditioning on
the parameter values from previous iteration \(\Theta^{(t-1)}\), \[
E_{\boldsymbol{\gamma}|\Theta^{(t-1)}}\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X}) .
\] Hereafter, we use the shorthand notation
\(E(\cdot)\equiv E_{\boldsymbol{\gamma}|\Theta^{(t-1)}}(\cdot)\). In the
M-step, we find the parameters \(\Theta^{(t)}\) that maximize
\(E\log f(\Theta, \boldsymbol{\gamma}| \textbf{y}, \textbf{X})\). The
parenthesized subscription \(^{(t)}\) denotes the parameter estimation
at the \(t\)th iteration. The E- and M- steps are iterated until the
algorithm converges.

@@

Our proposed model, BHAM, requires two preset scale parameters (\(s_0\),
\(s_1\)). Hence, we need to find the optimal values for the scale
parameters such that the model reaches its best prediction performance
regarding criteria of preference. This would be achieved by constructing
a two-dimensional grid, consisting of different (\(s_0\), \(s_1\))
pairs. However, previous research suggests the value of slab scale
\(s_1\) has less impact on the final model and is recommended to be set
as a generally large value, e.g.~\(s_1 = 1\), that provides no or weak
shrinkage. \citep{Rockova2018} As a result, we focus on examining
different values of spike scale \(s_0\). Instead of the two-dimensional
grid, we consider a sequence of \(L\) decreasing values
\(\{s_0^l\}: 0 < s_0^1 < s_0^2 < \dots < s_0^L < s_1\). Increasing the
spike scale \(s_0\) tends to include more non-zero coefficients in the
model. A measure of preference calculated with cross-validations (CV),
e.g.~deviance (defined as model log-likelihood times -2,
\(-2\log f(\boldsymbol{y}|\boldsymbol{\hat\beta}, \hat\phi)\)), area
under the curve (AUC), mean squared error, can be used to facilitate the
selection of a final model. The procedure is similar to the LASSO
implementation in the widely used R package \texttt{glmnet}, which
quickly fits LASSO models over a list of candidate values of \(\lambda\)
and gives a sequence of models for users to choose from.

@@

In this section, we apply the proposed model BHAM to analyze two
published metabolomics datasets where the outcomes are binary and
continuous respectively. We demonstrate the improved prediction
performance compared to the other Bayesian hierarchical additive model,
SB-GAM \citep{Bai2021}, while being computationally efficient (see Table
\ref{tab:time_real_data}). BHAM requires roughly 10\% of the computation
time of SB-GAM to fit models.
